### 2022-11-12 NSF2022/Description/specificaim2.tex

\subsection{Specific Aim 2: Efficient, high-order methods for
large-scale simulations}
\label{sec:specificaim2}
% ----------------------------------------------------------------------
Over the past decade, there has been an explosion of interest in
small-scale processes that utilize capillary forces and van der Waals
interactions to coordinate movement and bind microscopic components in
solvent~\cite{Pandey2011, Zhang2017, Siontorou2017}. Additionally,
hydrodynamic interactions are central to fabricate complex,
three-dimensional microstructures~\cite{Dasgupta2017, Leong2007,
Reynolds2019, Cho2010}. Hydrodynamic effects cannot be ignored since
they, like for the rates of biological functions like pore
dynamics~\cite{RYHAM20112929}, set the rate of dissipation. The success
of Specific Aims 1 and 3 depends on the ability to accurately and
efficiently simulate dense suspensions of granules governed by
equations~\eqref{eq:RBT}--\eqref{eqn:stressbalance}. Because of their
ability to resolve complex unbounded domains and achieve high-order
accuracy, integral equations (IEs) will be a central theme of the
computational tools.

To briefly describe IE formulations of PDEs, we consider the Stokes
equation~\eqref{eqn:stokes} as an example. Its solution can be written
as a convolution 
\begin{align*}
  \uu(\xx) = \int_{\partial\Omega(t)} G(\xx - \yy) \ssigma(\yy) \, ds_\yy,
\end{align*}
where $G(\xx - \yy)$ is a carefully chosen derivative of the PDE's
fundamental solution and $\ssigma: \partial \Omega(t) \rightarrow \RR^2$
is an unknown density function. The IE reformulation has several
immediate advantages including a dimension reduction since $\ssigma$ is
defined only on $\partial \Omega(t)$, and the far field boundary
condition of the PDE is automatically satisfied. The density
function $\ssigma$ is chosen to satisfy the boundary condition, and this
results in a IE 
\begin{align*}
  \gg(\xx) = \lambda \ssigma(\xx) + 
    \int_{\partial\Omega(t)} G(\xx - \yy) \ssigma(\yy) \, ds_\yy,
    \quad \xx \in \partial\Omega(t),
\end{align*}
where $\gg$ is the Dirichlet boundary condition, and $\lambda$ is a
constant that depends on the choice of $G$. For numerical stability,
choices of $G$ that result in second-kind Fredholm integral equations,
meaning $\lambda \neq 0$, are preferred.

\begin{wrapfigure}[27]{l}{1.8in}
  \vspace{-8pt}
  \centering
  \input{figures/doublewell.tikz} 
  \caption{\label{fig:CA} \footnotesize (a) A double well potential. (b)
  The solution of the one-dimensional phase
  transition~\eqref{eqn:phase}. (c) The convergence of exact and inexact
  Newton's methods.}
\end{wrapfigure}

The order parameter is governed by the PDE~\eqref{eqn:phase} in the
solvent phase $\Omega(t)$. For the single-well $W(\phi) = \phi^2$,
equation~\eqref{eqn:phase} is linear, and the PIs have used IEs to solve
for the order parameter~\cite{Fu2018_SIAM, FuQuRyYo22,
fu-ryh-qua-you2022}. However, for a double-well potential $W(\phi)$, the
resulting PDE is nonlinear, motivating the following problem:

% ----------------------------------------------------------------------
\begin{quotation}
  \noindent
  \textbf{Problem 4.} How can IEs be coupled with a Newton solver to
  solve the non-linear PDE~\eqref{eqn:phase}?
\end{quotation}

For a general double-well potential $W(\phi)$ with local minima at
$\phi_0$ and $\phi_1$, we propose to interpret the solution
of~\eqref{eqn:phase} as a root finding problem for
\begin{align}
  \label{eqn:phaseRoot}
  F[\phi] = -\epsilon^2 \Delta \phi + \tfrac{1}{2}W'(\phi).
\end{align}
Equation~\eqref{eqn:phaseRoot} will be solved with Newton's method,
$\phi^{N+1} = \phi^{N} - \alpha (dF^N)^{-1} F(\phi^N)$, where $dF^N =
-\epsilon^2 \Delta + \tfrac{1}{2}W''(\phi^N)$ is the Fr\'{e}chet
derivative of $F$, and $\alpha$ is the step size that will be determined
using backtracking. This requires solving the variable coefficient PDE
$-\epsilon^2 \Delta v + \tfrac{1}{2}W''(\phi^{N}) v = F[\phi^N]$, which
is challenging with an integral equation since a fundamental solution is
not readily available. However, by choosing a double-well potential with
$W''(\phi_0) = W''(\phi_1) = \beta$, we propose to instead use an
inexact Newton's method where $dF^{-1}$ is approximated with the
solution of $-\epsilon^2 \Delta v + \tfrac{\beta}{2} v = F[\phi^N]$. We
test the efficacy of the exact and inexact Newton's method applied
to~\eqref{eqn:phaseRoot} in $(0,1) \subset \RR$ by discretizing $dF$
with a second-order finite difference stencil. The double well potential
is illustrated in Figure~\ref{fig:CA}(a), and both the exact and inexact
Newton's methods converge to the solution in Figure~\ref{fig:CA}(b).
Figure~\ref{fig:CA}(c) shows the norm of the residual at each Newton
iterate when using the exact second derivative (red curve) and the
approximate second derivative (blue curve). While the inexact Newton's
method requires roughly twice as many iterations to reach of a tolerance
of $10^{-4}$, each iteration is significantly cheaper to apply.

@@

As discussed in Section~\ref{sec:bie},
numerical methods to solve IEs have several advantages over their
PDE-based counterpart. The PIs have demonstrated that IEs are a powerful
tool to simulate two-dimensional suspensions of amphiphilic
particles~\cite{Fu2018_SIAM, FuQuRyYo22, fu-ryh-qua-you2022}, and
elements of these formulations have been developed in three
dimensions~\cite{ying_2006, manasthesis, rac-gre2016}. In our current
implementation, we simulate $\mathcal{O}(100)$ two-dimensional granules,
and the proposed research will allow for the suspensions with
100X--1000X more two-dimensional granules and $\mathcal{O}(1000)$
three-dimensional granules.

% ----------------------------------------------------------------------
\begin{quotation}
  \noindent
  \textbf{Problem 5.} How can novel quadrature, time adaptivity, fast
  summation, and collision-detection be combined to accurately and
  efficiently solve the governing
  equations~\eqref{eq:RBT}--\eqref{eqn:stressbalance} in bounded and
  unbounded domains in two and three dimensions?
\end{quotation}

In experiments, it is known that interactions between bounding walls and
granules cannot be ignored~\cite{qua-gan-you2021,
KaouiHartingMisbah2011_PRE, her-sto-gra2005}. Therefore, the necessary
modification to the code will be performed so that confinement can be
considered. This modification requires introducing an additional layer
potential on the bounding walls, and elements of this modification have
already been incorporated (Figure~\ref{fig:stenosis}). However, to
consider rheological properties of the suspensions, a periodic domain,
such as a Couette device, will be considered. This multiply-connected
domain requires additional modifications. PI Quaife has considered
suspensions with confinement for several projects including dynamic
suspensions~\cite{qua-bir2014, bys-sha-qua2020} and porous
media~\cite{chi-moo-qua2020, dea-qua-bir-jua2018, moo-che-chi-qua2022,
qua-gan-you2021}.

The confinement introduces an additional computational
challenge---accurately computing integrals between nearly-touching
bodies. This issue is well-known in the IE community and several
algorithms have been developed in the last 30 years, including
panel-based methods~\cite{helsing_2008a, helsing_tutorial_2012},
Quadrature by Expansions (QBX)~\cite{din-hua-mar-tan2021,
klo-bar-gre-one2013, ste-bar2022, af2018sisc}, Barycentric-type formulae
for Cauchy integrals~\cite{bar2014, bar-wu-vee2015, ioa-pap-per1991},
and interpolation-based methods~\cite{ying_2006, qua-bir2014,
rah-vee-zor-bir2015} (Figure~\ref{fig:interpquad}). The proposed
research, which is described next, will combine several methods
including singularity subtraction, analytic function theory, and
spherical basis representations.

Since we are interested in round (2D) and spherical (3D) granules, we
will exploit the fact that integrals can be computed with spectral
accuracy by transforming the integrals to to a Fourier (2D) or spherical
harmonic (3D) basis. This technique can be used to evaluate layer
potentials for both Stokes~\cite{cor-vee2018} and the screened Laplace
equation~\cite{koh-cor-che-vee2021}. However, for other elliptical and
other shapes, an alternative approach is necessary. The Barycentric-type
formulae for Cauchy integrals results in a uniformly bounded quadrature
error, but the method is described only for the two-dimensional Laplace
and Stokes equations. PI Quaife recently showed that these quadrature
methods can be extended to other two-dimensional
PDEs~\cite{che-lin-her-qua2022}. The method uses singularity subtraction
write the fundamental solution of the PDE as a sum of two terms: a
logarithmic function and a continuously differentiable function. Then
the Barycentric-type formulae can be applied to the logarithmic
function, and the standard trapezoid rule~\cite{tre-wei2014} can be
applied to the other function. These singularity subtraction methods can
also accurately compute derivatives of layer potentials~\cite{car2021,
car2020, car-kha-kim2018} which is required, for example, when computing
the water-order stress~\eqref{eqn:stressbalance}. Quadrature is much
more challenging in three dimensions. Singularity subtraction methods
can be applied, but not combined with Barycentric-type formulae.
Alternatively, smoothed variations of the fundamental solution can be
used to address nearly-singular integrals~\cite{beale1, cor2001,
tlu-bea2019}.  Alternatively, efficient quadrature methods can be
applied directly to accurately compute singular and near-singular
integrals~\cite{khayat_2005, gimbutas2013sisc}.

While the inclusion of high-order quadrature and fast methods is
necessary to simulate dense suspensions, an additional reqired
computational tool is a collision-aware adaptive time stepping method.
The order parameter results in the granules attracting one another and
this leads to physical contact in finite time. Our current
implementation uses a repulsive Lennard-Jones force, but this introduces
stiffness, thereby requiring small time steps, and this can be a
bottleneck when simulating dense suspensions. Therefore, we will develop
a high-order collision-avoiding adaptive time stepping method. This
method will avoid collisions with by using the minimal separation
distance to impose a non-overlapping constraint. This results in a
linear complementarity problem that can be solve efficiently with a
projected gradient descent method~\cite{Yan2019}. High-order time
adaptivity will be based on PI Quaife's previous work on
suspensions~\cite{qua-bir2016}, and this method has proven to be a
robust method to simulate dense suspensions~\cite{qua-vee-you2019,
kab-qua-bir2017, qua-gan-you2021}. Following this previous work, the
proposed work will use a spectral deferred correction method since it
iteratively applies a low-order, single-step method to achieve
high-order accuracy. At each time step, we will compute the total force
and torque of the system which is computationally cheap and physically
zero, and therefore provides an estimate of the local truncation error.
PI Quaife's experience is that error estimates based on physical
constraints, such as force- and torque-free conditions, appropriately
determine adaptive time step sizes so that the dynamics are resolved
without the computational expense of techniques such as embedded
Runge-Kutta, step-doubling, and Richardson extrapolation.

>>>>

\subsection{Specific Aim 2: Efficient, high-order methods for
large-scale simulations}
\label{sec:specificaim2}
% ----------------------------------------------------------------------
Over the past decade, there has been an explosion of interest in
small-scale processes that utilize capillary forces and van der Waals
interactions to coordinate movement and bind microscopic components in
solvent~\cite{Pandey2011, Zhang2017, Siontorou2017}. Additionally,
hydrodynamic interactions are central to fabricate complex,
three-dimensional microstructures~\cite{Dasgupta2017, Leong2007,
Reynolds2019, Cho2010}. Hydrodynamic effects cannot be ignored since
they, like for the rates of biological functions like pore
dynamics~\cite{RYHAM20112929}, set the rate of dissipation. The success
of Specific Aims 1 and 3 depends on the ability to accurately and
efficiently simulate dense suspensions of granules governed by
equations~\eqref{eq:RBT}--\eqref{eqn:stressbalance}. Because of their
ability to resolve complex unbounded domains and achieve high-order
accuracy, integral equations (IEs) will be a central theme of the
computational framework.

To briefly describe IE formulations of PDEs, we consider the Stokes
equation~\eqref{eqn:stokes} as an example. Its solution can be written
as a convolution 
\begin{align*}
  \uu(\xx) = \int_{\partial\Omega(t)} G(\xx - \yy) \ssigma(\yy) \, ds_\yy,
\end{align*}
where $G(\xx - \yy)$ is a carefully chosen derivative of the PDE's
fundamental solution and $\ssigma: \partial \Omega(t) \rightarrow \RR^2$
is an unknown density function. The IE reformulation has several
immediate advantages including a dimension reduction since $\ssigma$ is
defined only on $\partial \Omega(t)$, and the far field boundary
condition of the PDE is automatically satisfied. The density
function $\ssigma$ is chosen to satisfy the boundary condition, and this
results in an IE 
\begin{align*}
  \gg(\xx) = \lambda \ssigma(\xx) + 
    \int_{\partial\Omega(t)} G(\xx - \yy) \ssigma(\yy) \, ds_\yy,
    \quad \xx \in \partial\Omega(t),
\end{align*}
where $\gg$ is the boundary condition, and $\lambda$ is a constant that
depends on the choice of $G$. For numerical stability, choices of $G$
that result in second-kind Fredholm integral equations, meaning $\lambda
\neq 0$, are preferred.

The PIs have demonstrated that IEs are a powerful tool to simulate
two-dimensional suspensions of amphiphilic particles~\cite{Fu2018_SIAM,
FuQuRyYo22, fu-ryh-qua-you2022}, and elements of these formulations have
been developed in three dimensions~\cite{ying_2006, manasthesis,
rac-gre2016}. In our current implementation, we simulate
$\mathcal{O}(100)$ two-dimensional granules, and the proposed research
will allow for the suspensions with 100X--1000X more two-dimensional
granules and $\mathcal{O}(1000)$ three-dimensional granules.

% ----------------------------------------------------------------------
\begin{quotation}
  \noindent
  \textbf{Problem 4.} How can novel quadrature, time adaptivity, fast
  summation, and collision-detection be combined to accurately and
  efficiently solve the governing
  equations~\eqref{eq:RBT}--\eqref{eqn:stressbalance} with dense
  suspensions of two- and three-dimensional granules in bounded and
  unbounded domains?
\end{quotation}

In experiments, it is known that interactions between bounding walls and
granules cannot be ignored~\cite{qua-gan-you2021,
KaouiHartingMisbah2011_PRE, her-sto-gra2005}. Therefore, the necessary
modification to the code will be implemented so that confinement can be
considered. This modification requires introducing an additional layer
potential on the bounding walls, and elements of this modification have
already been incorporated (Figure~\ref{fig:stenosis}). However, to
consider rheological properties of the suspensions, a periodic domain,
such as a Couette device, will be considered. Such multiply-connected
domains require additional modifications to the IE formulation. PI
Quaife has considered suspensions with confinement for several projects
including dynamic suspensions~\cite{qua-bir2014, bys-sha-qua2020} and
porous media~\cite{chi-moo-qua2020, dea-qua-bir-jua2018,
moo-che-chi-qua2022, qua-gan-you2021}.

The confinement introduces an additional computational
challenge---accurately computing integrals between nearly-touching
bodies. This issue is well-known in the IE community and several
algorithms have been developed in the last 30 years, including
panel-based methods~\cite{helsing_2008a, helsing_tutorial_2012},
Quadrature by Expansions (QBX)~\cite{din-hua-mar-tan2021,
klo-bar-gre-one2013, ste-bar2022, af2018sisc}, Barycentric-type formulae
for Cauchy integrals~\cite{bar2014, bar-wu-vee2015, ioa-pap-per1991},
and interpolation-based methods~\cite{ying_2006, qua-bir2014,
rah-vee-zor-bir2015} (Figure~\ref{fig:interpquad}). The proposed
research, which is described next, will combine several methods
including singularity subtraction, analytic function theory, and
analytic methods.

Since we are interested in round (2D) and spherical (3D) granules, we
will exploit the fact that integrals can be computed with spectral
accuracy by transforming the integrals to to a Fourier (2D) or spherical
harmonic (3D) basis. This technique can be used to evaluate layer
potentials for both Stokes~\cite{cor-vee2018} and the screened Laplace
equation~\cite{koh-cor-che-vee2021}. However, this quadrature is not
appropriate or other shapes, and an alternative approach is necessary.
The Barycentric-type formulae for Cauchy integrals results in a
uniformly bounded quadrature error, but the method is described only for
the two-dimensional Laplace and Stokes equations. PI Quaife recently
showed that these quadrature methods can be extended to other
two-dimensional PDEs~\cite{che-lin-her-qua2022}. The method uses
singularity subtraction write the fundamental solution of the PDE as a
sum of two terms: a logarithmic function and a continuously
differentiable function. Then the Barycentric-type formulae can be
applied to the logarithmic function, and the standard trapezoid
rule~\cite{tre-wei2014} can be applied to the other function. These
singularity subtraction methods can also accurately compute derivatives
of layer potentials~\cite{car2021, car2020, car-kha-kim2018} which is
required, for example, when computing the water-order
stress~\eqref{eqn:stressbalance}. Singularity subtraction methods can
also be applied in three dimensions, but not combined with
Barycentric-type formulae. Alternatively, smoothed variations of the
fundamental solution can be used to address nearly-singular
integrals~\cite{beale1, cor2001, tlu-bea2019}. Alternatively, efficient
quadrature methods can be applied directly to accurately compute
singular and near-singular integrals~\cite{khayat_2005,
gimbutas2013sisc}.

While the inclusion of high-order quadrature and fast methods is
necessary to simulate dense suspensions, an additional required
computational tool is a collision-aware adaptive time stepping method.
The order parameter results in the granules attracting one another and
this leads to physical contact in finite time. Our current
implementation uses a repulsive Lennard-Jones force, but this introduces
stiffness, thereby requiring small time steps, and this can be a
bottleneck when simulating dense suspensions. Therefore, we will develop
a high-order collision-avoiding adaptive time stepping method. This
method will avoid collisions with by using the minimal separation
distance to impose a non-overlapping constraint. This results in a
linear complementarity problem that can be solve efficiently with a
projected gradient descent method~\cite{Yan2019}. High-order time
adaptivity will be based on PI Quaife's previous work on
suspensions~\cite{qua-bir2016}, and this method has proven to be a
robust method to simulate dense suspensions~\cite{qua-vee-you2019,
kab-qua-bir2017, qua-gan-you2021}. Following this previous work, the
proposed work will use a spectral deferred correction method since it
iteratively applies a low-order, single-step method to achieve
high-order accuracy. At each time step, we will compute the total force
and torque of the system which is computationally cheap and physically
zero, and therefore provides an estimate of the local truncation error.
PI Quaife's experience is that error estimates based on physical
constraints, such as force- and torque-free conditions, appropriately
determine adaptive time step sizes so that the dynamics are resolved
without the computational expense of techniques such as embedded
Runge-Kutta, step-doubling, and Richardson extrapolation.

\begin{wrapfigure}[27]{l}{1.8in}
  \vspace{-8pt}
  \centering
  \input{figures/doublewell.tikz} 
  \caption{\label{fig:CA} \footnotesize (a) A double-well potential. (b)
  The solution of the one-dimensional phase
  transition~\eqref{eqn:phase}. (c) The convergence of exact and inexact
  Newton's methods.}
\end{wrapfigure}

The order parameter is governed by the PDE~\eqref{eqn:phase} in the
solvent phase $\Omega(t)$. If $W(\phi) = \phi^2$,
equation~\eqref{eqn:phase} is linear, and the PIs have used IEs to solve
for the order parameter~\cite{Fu2018_SIAM, FuQuRyYo22,
fu-ryh-qua-you2022}, but for a double-well potential $W(\phi)$, the
resulting PDE is nonlinear, motivating the following problem:

% ----------------------------------------------------------------------
\begin{quotation}
  \noindent
  \textbf{Problem 5.} How can IEs be coupled with a Newton solver to
  solve the non-linear PDE~\eqref{eqn:phase}?
\end{quotation}

For a general double-well potential $W(\phi)$ with local minima at
$\phi_0$ and $\phi_1$, we propose to interpret the solution
of~\eqref{eqn:phase} as a root finding problem for
\begin{align}
  \label{eqn:phaseRoot}
  F[\phi] = -\epsilon^2 \Delta \phi + \tfrac{1}{2}W'(\phi).
\end{align}
Equation~\eqref{eqn:phaseRoot} will be solved with Newton's method,
$\phi^{N+1} = \phi^{N} - \alpha (F'(\phi^N))^{-1} F(\phi^N)$, where
$F'(\phi^N) = -\epsilon^2 \Delta + \tfrac{1}{2}W''(\phi^N)$ is the
Fr\'{e}chet derivative of $F$, and $\alpha$ is the step size that will
be determined using backtracking. This requires solving the variable
coefficient PDE $-\epsilon^2 \Delta v + \tfrac{1}{2}W''(\phi^{N}) v =
F[\phi^N]$, which is challenging with an integral equation since a
fundamental solution is not readily available. However, by choosing a
double-well potential with $W''(\phi_0) = W''(\phi_1) = \beta$, we
propose to instead use an inexact Newton's method where $F^{-1}$ is
approximated with the solution of $-\epsilon^2 \Delta v +
\tfrac{\beta}{2} v = F[\phi^N]$. We test the efficacy of the exact and
inexact Newton's method applied to~\eqref{eqn:phaseRoot} in $(0,1)
\subset \RR$ by discretizing $F'$ with a second-order finite difference
stencil. The double-well potential is illustrated in
Figure~\ref{fig:CA}(a), and both the exact and inexact Newton's methods
converge to the solution in Figure~\ref{fig:CA}(b).
Figure~\ref{fig:CA}(c) shows the norm of the residual at each Newton
iterate when using the exact second derivative (red curve) and the
approximate second derivative (blue curve). While the inexact Newton's
method requires roughly twice as many iterations to reach of a tolerance
of $10^{-4}$, each iteration is significantly cheaper to apply.

@@
